% !TEX encoding = UTF-8 Unicode
% !TEX root = ../m2report.tex
%
We assume familiarity with propositional and predicate logic \cite{Huth:2004:LCS:975331},
automated theorem proving \cite{Fitting:1996:FLA:230183},
term rewriting \cite{Baader:1998:TR:280474},
decision procedures \cite{Kroening:2008:DPA:1391237},
and (maximum) satisfiability testing \cite{Biere:2009:HSV:1550723}.
Even so, for clarity we introduce
definitions and state lemmas and theorems without proofs in this section.
% to be used in the following sections.
Notions and notations follow (with deviations) the lecture notes \cite{AM2015tr} and \cite{GM2013ar}.
%____________________________________________________________________________________________________________
%\subsection{Signatures}\label{sec:signature}
%% DEF %%
\begin{definition}\label{def:signature}
A \coloremph{signature} $\mcF=\mcFfP$ is the disjoint union of \coloremph{function symbols} $\mcFf$
and \coloremph{predicate symbols} $\mcFP$.
The \coloremph{arity} of a symbol is the number of its arguments.
With $\mcF^{(i)}$ we denote symbols with arity $i$.
Function symbols $\mc\in\mcFf^{(0)}$ with no arguments are called \coloremph{constants}.
%Predicate symbols $\mP\in\mcFP^{(0)}$ with no arguments are propositions.
%The special binary equality symbol ${\foEQ}\in\mcFP$ is always available.
\end{definition}
%\begin{remark}
%We exclusively use the symbol $\sEQ$ for identity in this report.
%\end{remark}
\begin{definition}
An \coloremph{interpretation} $\mcI$ over a signature $\mcF$ consists
of a \coloremph{universe}, i.e.~non-empty set  $A$,
operations $\mf^\mcI: A^n\rightarrow A$  for every function symbol $\mf\in\mcFf$,
and possibly empty relations $\mP^\mcI\subseteq A^n$ for every predicate symbol $\mP\in\mcFP$.
A \coloremph{consistent interpretation} of the equational symbol ${\trEQ}\in\mcFP^{(2)}$ must respect the theory of equality
with reflexivity, symmetry, transitivity, function congruence, and predicate congruence.
\end{definition}
%____________________________________________________________________________________________________________
%\subsection{Terms}\label{sec:terms}
\begin{definition}
We build the set of \coloremph{(function) terms} $\mcTFV$
from a signature $\mcF$ and
a countable set of \coloremph{variables} $\mcV$ disjoint from $\mcF$.
Every variable $x\in\mcV$ is a term,
every constant $c\in\mcFf$ is a term, and
every expression $\mf(t_1,\ldots,t_n)$
for every function symbol $\mf\in\mcFf$ with arity $n>0$
and terms $t_1,\ldots,t_n$
is a term.
%
We build the set of general or \coloremph{predicate} terms $\mP(t_1,\ldots,t_n)$ from
predicate symbols $\mP\in\mcFP$ with arity $n$ and terms $t_1,\ldots,t_n$.
%
A term is \coloremph{ground} if its set of variables $\var(t)=\myemptyset$ is empty.
The set of all ground (function) terms is donated by $\mcT(\mcFf)$.
\end{definition}

%
%\begin{remark}
%Most of the following definitions are easily applicable on predicate terms too.
%\end{remark}
%%
%% DEF %%
\begin{definition}\label{def:position}
A \coloremph{position} is a finite sequence of positive integers.
The root position is the empty sequence $\epsilon$.
Position $pq$ is obtained by concatenation of positions $p$ and $q$.
%
The expression $t|_p$ denotes the subterm at position $p$
and $t[s]_p$ is obtained by replacing the subterm of $t$ at position $p$ with the term $s$.
%A (function) term is called \coloremph{linear} if it does not contain multiple occurrences of the same variable.
%
\end{definition}
%% DEF %%
%\begin{definition}\label{def:subterm}
%The~\coloremph{subterm} $t|_p$ of term $t$ \coloremph{at position} $p\in\pos(t)$ is defined as
%\begin{gather*}
%t|_p = \left\{\begin{array}{ll}
%		t 		& \text{if }p=\epsilon \\
%		t_i|_q	& \text{if }t=\mf(t_1,\ldots,t_n)\text{ and }p=iq
%	\end{array}\right.
%\end{gather*}
%%
%If $s$ is a term then $t[s]_p$ represents the term build by replacing the subterm of $t$ at position $p$ with the term $s$.
%\begin{gather*}
%t[s]_p = \left\{\begin{array}{ll}
%		s 		& \text{if }p=\epsilon \\
%		\mf(t_1,\ldots,t_i[s]_q,\ldots,t_n)	& \text{if }t=\mf(t_1,\ldots,t_n)\text{ and }p=iq
%	\end{array}\right.
%\end{gather*}
%A \coloremph{hole} denotes a special constant symbol $\ctxhole\in\mcFf^{(0)}$.
%A \coloremph{context} is a term $t$ with exactly one hole, i.e.~one occurrence $|t|_{\ctxhole}=1$.
%\end{definition}
%% DEF %%
\begin{definition}\label{def:substitution}
A \coloremph{substitution} is an assignment $\alpha$ from variables $x\in\mcV$ to terms in $\mcTFV$
with finite \coloremph{domain} $\dom(\alpha) = \{ x\in\mcV\mid\alpha(x) \neq x \}$.
We write substitutions as mappings $\sigma=\{ x_1\mapsto s_1,\ldots,x_n\mapsto s_n \}$
where $\dom(\sigma)=\{ x_1,\ldots,x_n \}$ and $\sigma(x_i)=s_i$.
A \coloremph{variable substitution} is an assignment from $\mcV$ to $\mcV$.
A \coloremph{renaming} is a bijective variable substitution.
A \coloremph{proper instantiator} is a substitution that is not a variable substitution.
%We define the \coloremph{instance} $t\sigma$ of a term $t$
%\begin{align*}
%	t\sigma &= \left\{\begin{array}{ll}
%		s_i 					& \text{if }t=x_i\in\dom(\sigma)\\
%		t					& \text{if }t\in\mcV\,\backslash\dom(\sigma)\\
%		\mf(t_1\sigma,\ldots,t_n\sigma)	&\text{if }t=\mf(t_1,\ldots,t_n)
%	\end{array}\right.
%\end{align*}
%and the \coloremph{composition} of two substitutions $\sigma$ and $\tau$ as
%\begin{align*}
%	\sigma\tau&=\{ x_i\mapsto s_i\tau\mid x_i\in\dom(\sigma) \}
%	\cup
%	\{ y_i\mapsto t_i\mid y_i\in\dom(\tau) \backslash \dom(\sigma) \}.
%\end{align*}
\end{definition}
%% DEF %%
\begin{definition}\label{def:unifier}
Two terms $s$ and $t$ are \coloremph{unifiable} if there exists a substitution such that $s\sigma=t\sigma$.
They are \coloremph{variants} if their most general unifier is a renaming.
The \coloremph{most general unifier} $\sigma=\mgu(s,t)$ is a unifier such that
for every other unifier $\sigma'$ there exists a substitution $\tau$ such that
$\sigma' = \sigma \tau$.
\end{definition}
%% DEF %%
%\begin{example}
%The unifier $\sigma'=\{ x\mapsto\mg(\ma),y\mapsto\mb,z\mapsto\mg(\mb)\}$
%of terms $s=\mf(x,\mg(y))$
%and  $t=\mf(\mg(\ma),z)$
%is less general than $\sigma=\mgu(s,t)=\{ x\mapsto\mg(\ma), z\mapsto\mg(y)$:
%$\sigma'=\sigma\{ y\mapsto\mb \}$.
%\end{example}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{First order formulae}
%
\begin{definition}\label{def:first-order}
%Let $\mcF$ be a signature.
A \coloremph{first order formula} $F$ is either an
\coloremph{atomic formula}, i.e.~a predicate term or an equation, or a composition of formulae where $\mP$ is a predicate symbol and $s,t,t_1,\ldots,t_n$ are terms.
%
\[
 F ::= \mP(t_1,\ldots,t_n) \mid
	s \foEQ t \mid
	\lnot F \mid
	\forall x F \mid
	\exists x F \mid
	 F\lor F \mid
	 F\land F \mid
	 F\rightarrow F
\]
%negation $\lnot F$,
%universal quantification $\forall x F$,
%existential quantification $\exists x F$,
%%of one formula $\phi$;
%disjunction $ F\lor F'$,
%conjunction $ F\land F'$,
%or implication $ F\rightarrow F'$.
%Exclusive disjunction $ F\oplus F$ abbreviates $( F\lor F)\land(\lnot\phi\lor\lnot F)$.
%of two formulae $\phi$ and $\psi$.
\end{definition}
%% REM %%
%% DEF %%
\begin{definition}\label{def:literals}
A \coloremph{literal} $L$ is either an atom
or the negation of an atom.
%, where $l\not{}\hspace{-0.5mm}\circ$ abbreviates $\lnot(l\circ r)$
%for binary symobl $\circ$.
%A literal is ground if all its terms are ground.
%The complement $L^c$ of a literal $L$ is defined as %with some atom $A$:
%\[ L^c = \left\{ \begin{array}{rl}
%A & \text{if } L = \lnot A\\
%\lnot A & \text{if } L = A
%\end{array}\right.
%\]
Two literals are complementary if one is an atom and the other is the negation of this atom.
%
A \coloremph{clause}\ \ $\mcC = L_1\lor\ldots\lor L_n$  is a possible empty multiset of literals and is interpreted as disjunction of literals.
The \coloremph{empty clause} $\emptyclause$ expresses a contradiction.
%A clause is ground if all its literals are ground.
A finite \coloremph{set of clauses} $S=\{ \mcC_1,\ldots,\mcC_n \}$ 
is interpreted as an universally quantified conjunction of variable distinct clauses
$S\equiv\forall\vec{x}_1\ldots\forall\vec{x}_n\left(\mcC_1\land\ldots\land\mcC_n\right)$ with
$\var(\mcC_i)=\vec{x}_i$,
$\vec{x}_i\cap\vec{x}_j=\myemptyset\lor i=j$.
%A set of clauses is ground if all its clauses are ground.
\end{definition}
%
\begin{remark}
When the same variable symbols are used in different clauses the distinguishing additions, e.g.~super- or subscripts, were omitted due to laziness.
\end{remark}
%
%\begin{definition}
%A {\em clause} $\mathcal C = L_1\lor \ldots\lor L_n$ is a possible empty multiset of literals and is interpreted as disjunction of literals.
%$\square$ denotes the {\em empty clause}, respectively false, contradiction, pair or conjunction of complementary literals.
%A clause is ground if all its literals are ground.
%A finite {\em set of clauses} $S=\{ \mathcal C_1,\ldots \mathcal C_n \}$ is interpreted as universally quantified conjunction of all its clauses.
%\end{definition}
%
\begin{definition}\label{def:model}
%A ground literal $L$ holds in an Herbrand \coloremph{interpretation} $\mathcal I$ (a domain of ground terms and definitions for functions and predicates),
%if it is the logical consequence $\mathcal I\models L$ of ground literals in $\mathcal I$.
%A non-ground literal holds in $\mathcal I$, if all its ground instances hold in $\mathcal I$.
%A clause holds in $\mathcal I$, if at least one of its literals holds in $\mathcal I$.
%A set of clauses $S$ holds in $\mathcal I$, if every clause $\mathcal C\in S$ holds in $\mathcal I$.
%An interpretation is consistent $\mathcal I\not\models\square$, if no pair of complementary literals holds in $\mathcal I$.
%A set of clauses $S$ is \coloremph{satisfiable} if it has a \coloremph{model} $\mathcal M$, i.e.~a consistent interpretation $\mathcal M\models S$.
A \coloremph{term model} $\mcM$ is an consistent interpretation where the elements of the domain $A$
are equivalence classes of ground terms
and the interpretation of each ground term $t^\mcM := [t]_\sim$ is its equivalence class.
An equation $s\trEQ t$ of ground terms holds in $\mcM$ if $[s]_\sim=[t]_\sim$;
a ground predicate term $\mP(t_1,\ldots,t_n)$ holds in $\mcM$ if $\mP([t_1]_\sim,\ldots,[t_n]_\sim)] \in A^n$.
A ground literal holds in $\mcM$ if and only if its complementary literal does not hold in $\mcM$.

A non-ground literal holds in $\mcM$ if all its
%(infinitely many in the presence of function symbols with arity greater than one)
(perhaps infinitely many)
instances hold in $\mcM$.
A clause holds in $\mcM$, if at least one of its literals holds in $\mcI$.
A set of clauses $S$ holds in $\mcM$, if every clause $\mcC\in S$ holds in $\mcM$.
Then we say $\mcM$ is a model {\em for} $S$.

\end{definition}
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% SUBSEC %%

%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Term rewrite systems}
%% DEF %%
\begin{definition}
A \coloremph{rewrite rule}  $l \rightarrow r$ is an equation of terms where the left-hand side is not a variable
and the variables which occur in the right-hand side occur also in the left-hand side.
%\[
%	l\not\in\mcV\text{ and }\var(r)\subseteq\var(l)
%\]
A rewrite rule $l'\rightarrow r'$ is a \coloremph{variant} of $l\rightarrow r$ if there is a renaming $\varrho$ such that
$(l\rightarrow r)\varrho = (l\varrho\rightarrow r\varrho) = l'\rightarrow r'$.
A \coloremph{term rewrite system} is a set of rewrite rules without variants.
%\end{definition}
%% REM %%
%\begin{remark}
%It's easy to see (lemma \vref{lem:xa}) that not all equations $s\foEQ t$ or their inverse $t\foEQ s$ are rewrite rules.
%%A little sloppy we may write $l\trEQ r$ to express that $l\rightarrow r$ or its inverse $l\leftarrow r$ must be rewrite rules,
%%e.g.~$x\trEQ f(x)$.
%%If we do not care we write $l\foEQ r$.
%\end{remark}
%
%\begin{definition}
The expression $s\rightarrow t$ denotes a \coloremph{rewrite step}
from term $s$ to term $t$
if there exists
a position $p\in\pos(s)$,
a rewrite rule $l\rightarrow r$,
and a substitution $\sigma$ such that
$s|_p=l\sigma$ and $t=s[r\sigma]_p$.
The subterm $l\sigma$ is called \coloremph{redex} and
$s$ rewrites to $t$ by \coloremph{contracting} $l\sigma$ to \coloremph{contractum} $r\sigma$.
%
We may write
$s\rightarrow_{p,l\rightarrow r,\sigma}t$ to indicate position, rewrite rule and substitution
or $s\rightarrow_\mcR t$ to indicate the origin of the rule.
%
A term $s$ is \coloremph{irreducible} or in \coloremph{normal form} if there is no rewrite step $s\rightarrow t$ for any term $t$.
The set of normal forms $\mNFR$ of a TRS $\mcR$ contains all irreducible terms of the TRS.
%
\end{definition}
%%%
\begin{definition}
A term $s$ can be rewritten to term $t$ with notion $s\rightarrow^* t$
if there exists at least one \coloremph{rewrite sequence} $(a_1,\ldots ,a_n)$ such that
$s=a_1$, $a_n=t$, and $a_i\rightarrow a_{i+1}$ are rewrite steps for $1\leq i<n$.
A TRS is \coloremph{terminating} if there is no infinite rewrite sequence of terms.
%
Two Terms $s$ and $t$ are \coloremph{joinable} with notion $s\downarrow t$
if both can be rewritten to some term $c$, i.e.~$s \rightarrow^*c\ \, ^*\!\!\leftarrow t$.
A TRS is \coloremph{confluent } if $s$ and $t$ are joinable whenever $s\ ^*\!\!\leftarrow a \rightarrow^* t$ holds for some term $a$.
%
Terms $s$ and $t$ are \coloremph{convertible} with notion $s\leftrightarrow^* t$
if there exists a sequence $(a_1,\ldots ,a_n)$ such that
$s=a_1$, $a_n=t$, and $a_i\leftrightarrow a_{i+1}$, i.e.~$a_i\rightarrow a_{i+1}$ or $a_i\leftarrow a_{i+1}$ are rewrite steps for $1\leq i<n$.
\end{definition}
%
%% DEF %%
\begin{definition}\label{def:closed-under}
%A binary relation $R$ on terms is \coloremph{closed under contexts}
%if $u[s]|_p\,R\;u[t]|_p$ holds
%for all positions $p\in\pos(u)$ and
%for all terms $s,t,u$ with $s\,R\;t$.
%It is \coloremph{closed under substitutions}
%if $s\sigma\,R\;t\sigma$ holds for all substitutions $\sigma$
%and all terms $s,t$ with $s\,R\;t$.
A binary relation $\relation$ on terms is called a \coloremph{rewrite relation} whenever it is \coloremph{closed under contexts},
i.e.~$u[s]|_p\relation u[t]|_p$ %holds
for all positions $p\in\pos(u)$ and
for all terms $s,t,u$ with $s\relation t$
and \coloremph{closed under substitutions},
i.e.~$s\sigma\relation t\sigma$ %holds
for all substitutions $\sigma$
and all terms $s,t$ with $s\relation t$.
\end{definition}
\begin{lemma}
The relations $\rightarrow^*_\mcR$,
$\rightarrow^+_\mcR$,
$\downarrow_\mcR$, $\uparrow_\mcR$ are rewrite relations on every TRS $\mcR$.
\end{lemma}
%
\begin{definition}
A proper (i.e.~irreflexive and transitive) order on terms is called \coloremph{rewrite order} if it is a rewrite relation.
A \coloremph{reduction order} is a well-founded rewrite order,
i.e.~there is no infinite sequence
$(a_i)_{i\in\mathbb{N}}$
where $a_i\gtpre a_{i+1}$ for all $i$.
% with $i\in\mathbb{N}$.
A \coloremph{simplification order} is a rewrite order with the \coloremph{subterm property},
i.e.~$u[t]_p \gtpre t$ for all terms $u$, $t$ and positions $p\neq\epsilon$.
\end{definition}
\begin{lemma}
Every simplification order is well-founded, i.e.~it is a reduction order.
\end{lemma}
%% DEF %%
%
\begin{theorem}
A TRS $\mcR$ is terminating if and only if there exists a reduction order $\gtpre$
such that $l\gtpre r$ for every rewrite rule $l\rightarrow r\in\mcR$.
We call $\mcR$ simply terminating if $\gtpre$ is a simplification order.
\end{theorem}

%\begin{theorem}[Knuth-Bendix-Criterion]
%\end{theorem}
%% DEF %%

