% !TeX root = ../mythesis.tex
% !TeX encoding = UTF-8
% !TeX spellcheck = en_US

% In the following definitions \( \mcf \) denotes a function symbol, a predicate symbol or the equality sign.


\begin{definition}\label{def:substitution}
	A \coloremph{substitution} \( \sigma \) is a mapping from variables \( x\in\mcV \) to terms in \( \mcTFfV \)
	where \coloremph{domain }\( \domain(\sigma) = \{ x\in\mcV\mid\sigma(x) \neq x \} \)
	and image \( \image(\sigma) =\{ \sigma(x) \mid x\in V, \sigma(x)\neq x \} \) are finite.
	We write substitutions as bindings, e.g.~\( \sigma=\{ x_1\mapsto s_1,\ldots,x_n\mapsto s_n \} \)
	where \( \domain(\sigma)=\{ x_1,\ldots,x_n \} \) and \( \sigma(x_i)=s_i \).
	A \coloremph{variable substitution} is a mapping from \( \mcV \) to \( \mcV\subseteq\mcTFfV \).
	A \coloremph{renaming} is a bijective variable substitution.
	A \coloremph{proper instantiator} is a substitution that is not a variable substitution
	(at least one variable is mapped to a non-variable term).
\end{definition}
\begin{definition}
	We define the application of a substitution \( \sigma \)
	to term \( t \), a literal \( L \), or a clause \( C \) as follows
	\begin{align*}
	t\sigma &=
	\left \{
		\begin{array}{ll}
			s_i & \text{if }t=x_i\in\domain(\sigma), \sigma(x_i) = s_i
			\\
			y & \text{if }t=y\in\mcV \, \backslash\domain(\sigma)
			\\
			\mf(t_1\sigma,\ldots,t_n\sigma)	&\text{if }t=\mf(t_1,\ldots,t_n)
			\text{ where  }\mf\in\mcFfn
		\end{array}
	\right.
			 \\
	L\sigma{} &=
	\left \{
		\begin{array}{ll}
			\mP(t_1\sigma,\ldots,t_n\sigma)
			&\text{if }L=\mP(t_1,\ldots,t_n)
			\text{ where  }\mP\in\mcFPn
			\\
			t_1\sigma\mEQ t_2\sigma
			&\text{if } L = t_1\mEQ t_2
			\\
			\lnot(A\sigma)&\text{if } L=\lnot A\text{ where }A\text{ is an atom}
		\end{array}
	\right.
	\\
	C \sigma{} &=
\bigvee_{
	L\in{} C
}
	L \sigma{}
	\end{align*}
\end{definition}

\begin{definition}\label{def:term:literal:clause:substitution}
	We define the application of special grounding “substitution” \( \bot \) to a term \( t \) with distinct constant symbol \( \bot \not\in\mcF(t) \)
	as follows
	\begin{gather*}
	\MDEFINE{t\sigma}{ll}{
			\bot & \text{if }t=x\in\mcV
			\\
			\mf(t_1\sigma,\ldots,t_n\sigma)	&\text{if }t=\mf(t_1,\ldots,t_n)
			\text{ where  }
			 \mf\in\mcFfn
			 }
	\end{gather*}
	We define the application of \( \bot \) to literals and clauses as in
	Definition~\vref{def:term:literal:clause:substitution}.

\end{definition}

\begin{definition}
	We can easily extend our definition to composite first order formulae,
	but the cases of quantified formulae would need more consideration
	and are not needed in this thesis.
	So we only partially define \( F\sigma \) for first order formulae \( F \) and substitution \( \sigma \) as follows (if \( G\sigma \) and \( H\sigma \) are defined in the respective cases).
	\begin{gather*}
	\MDEFINE{F\sigma}{ll}{
		\lnot(G\sigma) & \text{if }F = \lnot G
		\\
		(G\sigma) * (H\sigma) & \text{if }F = G * H, *\in \{ \land, \lor, \limp \}
		\\
		\quantify x (G\sigma) &\text{if }F = \quantify x G,
		\quantify \in \{ \forall, \exists \},
		x\not\in\domain(\sigma)
		% \\
		% \texttt{undefind} & \texttt{otherwise}
		% x\not\in\bvar(G)
	}
	\end{gather*}
\end{definition}

\begin{definition}
	A clause \( \mcC \) \coloremph{strictly subsumes} a clause \( \mcD \) if their exists a substitution \( \theta \) such that \( \mcC\theta \subsetneq \mcD \),
	e.g.~when clause \( \mcD = \mcC\theta \lor \mcD' \) is a weakened instance of clause \( \mcC \).
\end{definition}
%\begin{example}
%	Trivially any clause \( \mcC \) subsumes all its instances \( \mcC\sigma \) or weakenings \( \mcC\lor\mcD \).
%	Further \( x\mEQ y \) subsumes \( x\mEQ x \lor \mP(y) \).
%\end{example}

\begin{definition}\label{def:substcomp}
We define the \coloremph{composition} of two substitutions \( \sigma \) and \( \tau \) as follows
	\begin{align*}
		\sigma\tau&=\{ x_i\mapsto s_i\tau\mid x_i\in\domain(\sigma) \}
		\cup
		\{ y_i\mapsto t_i\mid y_i\in\domain(\tau) \backslash \domain(\sigma) \}.
	\end{align*}
\end{definition}

\begin{lemma}\label{lem:substitution}
	With the definitions in~\ref{def:substitution} and~\ref{def:substcomp} the equation
	\( (\mct\sigma)\tau = \mct(\sigma\tau) \) holds for
	term, atoms, and literals.
\end{lemma}

\begin{proof}
	Assume \( \sigma \) and \( \tau \) are substitutions.
	Then we use induction on the structure of the 	expression \( \mct \)
	that the equation \( (\mct\sigma)\tau =  \mct(\tau\sigma) \) holds in all possible cases.
	\begin{itemize}
		\item (base case) Let \( \mct = x_i \in\domain(\sigma) \) then
		\( ((x_i)\sigma)\tau\defEQ s_i\tau\defEQ x_i(\sigma\tau) \) holds.

		\item (base case) Let \( \mct = y \not\in\domain(\sigma) \) then
		\( (y\sigma)\tau \defEQ y\tau \defEQ y(\sigma\tau) \) holds.

		\item (step case) Let \( \mct = \mcf(t_1,\ldots,t_n) \)
		then
		\(
			((\mcf(t_1,\ldots,t_n))\sigma)\tau
		\defEQ
		(\mcf(t_1\sigma,\ldots,t_n\sigma))\tau
		\defEQ
		\mcf((t_1\sigma)\tau,\ldots,(t_n\sigma)\tau)
		\defEQ[IH]
		\mcf(t_1(\sigma\tau),\ldots,t_n(\sigma\tau))
		\defEQ
		(\mcf(t_1,\ldots,t_n))(\sigma\tau)
		\) holds.

		\item (step case) Let \( \mct = \lnot A \) then
		\(
		((\lnot A)\sigma)\tau
		\defEQ
		(\lnot(A\sigma))\tau
		\defEQ
		\lnot((A\sigma)\tau)
		\defEQ[IH]
		\lnot(A(\sigma\tau))
		\defEQ
		(\lnot A)(\sigma\tau)
		\) holds.
	\end{itemize}
\end{proof}
