% !TeX root = ../mthesis.tex
% !TeX encoding = UTF-8
% !TeX spellcheck = en_US




\chapter{Mathematics}

\input{epigraphs/Menge}



In this chapter we recall some mathematical and logical concepts and notions 
we have used but not defined in the main part.
We assume at least a {\myem basic} understanding of sets as already 
described by Georg Cantor in the 19th century. 
However we just assume that our sets are definable within a consistent set theory.
Thus we have well defined {\myem membership} relationships between objects and sets,
well defined {\myem subset} relationships between sets, 
and well defined {\myem Cartesian products} over sets.

In section \ref{sec:app:maths} we {\em finally} define {\myem relations} and {\myem orders}. 
In section \ref{sec:app:fol} we describe sets and numbers general calculated from first order terms.



\section{Relations and orders}\label{sec:app:maths}

\input{appendix/math}

\section{First order logic}\label{sec:app:fol}

\input{appendix/termsets}


\chapter{Conventions}

\begin{figure}[hbt]
	\begin{align*}
	x,y,z &\ \in\mcV \tag*{variables} \\
	\ma,\mb,\mc,\md &\ \in\mcFfn[{0}]\tag*{constant symbols} \\
	\mf, \mg, \mh &\ \in\mcFfn[{n>0}]\tag*{function symbols} \\
	\mpp, \mq, \mr &\ \in\mcFPn[{0}]\tag*{propositional symbols}\\
	\mP, \mQ, \mR &\ \in\mcFPn[{n>0}]\tag*{predicate symbols}\\
	\mcf, \mcg, \mch &\ \in\mcF^{n>0} \tag*{predicate or functions symbols}\\
	\ell, r, s,t,u &\ \in\mcTf = \mcTFfV\tag*{(first order) terms}\\
	\mcs, \mct, \mcu &\ \in \mcPT \cup \mcET \cup \mcTf \tag*{predicates, equations, terms}\\
	\mu, \nu, \sigma, \tau &\ \tag*{substitutions, unifiers}\\
	A,B,C,D &\ \tag*{atoms} \\
	\mcL, L &\ \tag*{literals} \\
	%	\mcC, \mcD &\ \tag*{clauses} \\
	S &\ \tag*{sets of clauses}
	\end{align*}
	\caption{Typographical conventions}
	\label{fig:conventions}
\end{figure}

\section{First Order Formulae}

In Section \ref{sec:syntax} we have defined atomic formulae in Definition \ref{def:atoms}, 
but we can only build formulae in clausal normal form (\CNF) with Definition \vref{def:CNF}.
Now we will define arbitrary first order formulae (\FOF).

\begin{definition}[\FOF]
	Predicates and equations are (atomic) first order formulae. 
	The negation $(\lnot F)$, 
	the universal quantification $(\forall x F)$, 
	and the existential quantification $(\exists x F)$ 
	of a given formula $F$ are (composite) first order formulae.
	Further, the disjunction $(F \lor F')$, 
	the conjunction $(F \land F') $, 
	and the implication $(F \limp F') $ 
	of two given formulae $F$ and $F'$ 
	are (composite) first order formulae.
\end{definition}

We've already defined when an atom holds for an assignment $\alpha_\mcI$ 
in an interpretation $\mcI$ within Definition \vref{def:model}.
Now we extend these definitions to arbitrary formulae.

\begin{definition}[Semantics of \FOF]
	
	A universally quantified formula $\forall x F$ holds in $\mcI$ if its subformula $F$ holds for all assignments for $x$.
	An existential quantified formula $\exists xF$ holds if its subformula $F$ holds for at least one assignment for $x$.
	A negation $\lnot F$ holds if its subformula $F$ does not hold, 
	a disjunction $F\lor F'$ holds if one or both of its subformulae $F$ or $F'$ hold,
	a conjunction $F\land F'$ holds, if both of its subformualae $F$ and $F'$ hold, 
	an implication $F\limp F'$ holds if its first subformula $F$ does not hold or its second subformula $F'$ holds (or both).
\end{definition}

\begin{remark}Usually we us precedences on connectives to omit parentheses 
	and some heuristics to structure the formulae for readability 
	without introducing semantic ambiguity.
%
	Beside the obvious semantically indistinguishable formulae with double negations, conjunctions, and disjunctions 
	we have introduced new ones.
	\begin{enumerate}
		\item $\forall x F$, $\exists x F$, and $F$ are indistinguishable if $x\not\in\var(F)$. 
		We usually omit quantifiers with variables that do not occur in subformulae.
		\item In general $\exists x F$ is different from $F$ if $x\in\var(F)$, e.g. $\exists x(x\mNE\ma)$ is satisfiable and $x\mNE\ma$ isn't.
		\item $\forall x F$ and $F$ are equivalent even if $x\in\var(F)$, 
		because in both cases we demand that $F$ holds in all assignments in our model.
		Usually we keep these universal quantifiers in \FOF.
		
		A first order formulae without quantifiers is in {\myem clausal form}, 
		but not necessarily in \CNF, e.g.~a weakened version of symmetry $(x\mEQ \ma)\limp (\ma\mEQ x)$ 
		is equisatisfiable to $\forall x ((x\mEQ \ma)\limp (\ma\mEQ x))$ 
		or $\exists a (\forall x ((x\mEQ a)\limp (a\mEQ x))$. 
	\end{enumerate}

\end{remark}

%\subsection{Clause Normal Form}












