% !TeX root = ../mythesis.tex
% !TeX encoding = UTF-8
% !TeX spellcheck = en_US


\chapter{Undecidability}

A logical calculus, i.e.~a formal (or purely syntactical) proof system for an underlying logic,
is \coloremph{complete} if every (semantically) valid formula is (syntactically) 
provable from its premises by the calculus.
In other words, every sentence that holds in all possible models for its premises
is derivable from its premises by applying rules of the formal system only.
Naturally we expect a useful calculus to be \coloremph{sound}, that is,
every (syntactically) provable formula (semantically) holds in any model for its premises.
Without premises a sentence has to be provable if and only if it holds in any interpretation of its signature.

% A logic or theory is \coloremph{decidable} if a decision procedure exists,
% i.e.an algorithm that eventually terminates with the correct answer.


%A standard example for a decidable logic is propositional logic.

We first state some completeness, undecidability, and other
fundamental theorems about first order logic
in Section~\vref{sec:fol:theorems}.
Then we enumerate decidable fragments of first order logic
which can be described purely syntactically
in Section~\vref{sec:decidable:fol:fragments}.
We conclude this chapter with a look at decidable first-order theories
in Section~\vref{sec:decidable:fol:theories},
which are not necessarily contained
in one of the syntactically describable and decidable fragments of first-order logic.


\section{Theorems about First Order Logic}\label{sec:fol:theorems}

The fundamental theorems about first order logic
were already proven in the first half of the 20th century.
They outline the possibilities and limitations of any attempt
to create a general procedure that checks the validity of arbitrary first order sentences.

\begin{theorem}[Soundness]
	The inference rules of natural deduction (see Definition\vref{def:natural:deduction}) are sound.
\end{theorem}

\begin{proof}
	We can prove the soundness of each inference rule by case distinction and the use of the semantic definition of validity.
\end{proof}

\input{theorems/goedelsvollstaendigkeitssatz}

\begin{theorem}[\cite{bennett1973}]
	Natural deduction is a complete calculus,
	i.e.\,a proof \( \Gamma\proves G \) exists,
	whenever \( \Gamma\models G \).
\end{theorem}

% \begin{theorem}[Completeness, Gödel 1929]
% 	Natural deduction --- a sound formal proof system for first order logic --- is complete.
% \end{theorem}



\begin{lemma}[Refutation]
	A formula is valid if and only if its negation is not satisfiable.
\end{lemma}
\begin{proof}
By definition of the semantics of negation, validity, and satisfiability.
\end{proof}

\begin{theorem}
	[Undecidability~\cite{church_1936,turing1937}, Church 1936,
	Turing 1937.]
	The satisfiability problem for first order logic is undecidable.
\end{theorem}

\begin{theorem}[Trakhtenbort 1950, Craig 1950]
	The satisfiability problem for first order logic on \coloremph{finite} structures (domains) is undecidable.
\end{theorem}

\begin{definition}[Finite model property]
	A logic has the finite model property if each non-theorem is falsified by some finite model.
\end{definition}

\begin{theorem}[Compactness, Gödel 1930, Maltsev 1936]\label{the:compactness}
	If every finite subset of a set of formulas \( S \) has a model then \( S \) has a model.
\end{theorem}

\begin{theorem}[Löwenheim-Skolem, 1915, 1920]\label{the:loewenheim}
	If a set of formulas \( S \) has a model then \( S \) has a countable model.
\end{theorem}

\begin{theorem}[Herbrand, 1930]\label{the:herbrand}
	Let \( S \) be a set of clauses without equality. Then the following statements are equivalent.
	\begin{itemize}
		\item \( S \) is satisfiable.
		\item \( S \) has a Herbrand model.
		\item Every finite subset of all ground instances of \( S \) has a Herbrand model.
	\end{itemize}
\end{theorem}

\begin{corollary}
	Let \( S \) be a set of clauses without equality.
	Then \( S \) is unsatisfiable if and only if there exists
	an unsatisfiable finite set of ground instances of \( S \).
\end{corollary}

\begin{lemma}
	Skolemization preserves satisfiability.
\end{lemma}

\begin{lemma}
	Tseitin transformation preserves satisfiability.
\end{lemma}

% \begin{lemma}
% 	Herbrandization preserves validity.
% \end{lemma}

\begin{lemma}
With skolemization, Herbrandization, and Tseitin transformation introduced 
we can effectively transform an arbitrary first-order formula 
into an equisatisfiable set of clauses.
\end{lemma}


\section{Decidable Fragments of First Order Logic}\label{sec:decidable:fol:fragments}

In first order logic validity and satisfiability are undecidable for arbitrary formulae.
Even so this section presents purely syntactically defined fragments
of first order logic where satisfiability is decidable.%\footnote{
% 	\,Definitions and compact overviews follow the presentation
% 	“\href{http://logic.rwth-aachen.de/~graedel/kalmar.pdf}{Decidable fragments of first-order and fixed-point logic}”
% 	by E.~Grädel (\url{http://logic.rwth-aachen.de/~graedel/}).
% }

\begin{definition}[Prefix-vocabulary classes \cite{MR1482227}]\label{def:prefix:class}
	We introduce a simple notation to define fragments 
	of first order logic---i.e.~sets of first order sentences in \PNF{}---by 
	the syntactic structure of the included sentences.
	\[
		{[ \, \Pi, (p_i)_{i>0}, (f_i)_{i>0} \, ]}_{(\mEQ)}
		\subseteq {[\, all, all, all\,]}_{\mEQ}
	\]
	where 
	%      
	\begin{enumerate}
		\item \( \Pi \) is an expression over quantifier symbols 
		\( \{ \forall\,\exists \} \) 
		to define the set of possible prefixes of sentences in the fragment, 
		e.g.~with \( \Pi = \forall^*\exists \) we have 
		\( \mcL(\Pi) = \{ \exists, \forall \exists, \forall\forall\exists, \ldots \} \)
		where prefix \( \forall x \exists y \) matches,
		but \( \exists x \exists y \) does not.
	% 
	\item \( (p_i)_{i>0} \) is an infinite sequence of 
	\( p_i \in \mathbb{N}\disjointunion\{ \omega \} \) where 
	\( p_i \) represents the maximal number of predicate symbols 
	with arity \( i \in \mathbb{N}\disjointunion\{ \omega \} \),
	where \( \omega \) allows a countable set of symbols,
	% 
	\item \( (f_i)_{i\in\mathbb{N}} \) is an infinite sequence of 
	\( f_i \in \mathbb{N}\disjointunion\{ \omega \} \) where
	\( f_i \) represents the maximal number of function symbols with arity \( i \),
	% 
	\item the equality symbol is not counted as binary predicate symbol. 
	Instead, the absence or presence of equality in the fragment
	is indicated by the absence or presence of a subscript \( \mEQ \),
% 
	\item and infinite sequences \( (v_i)_{i>0} \) are represented by
	finite sequences \( (v_1,v_2,\ldots,v_\ell) \) if
	\( v_j = 0 \) for all \( j > \ell \).
\end{enumerate}
\begin{remark}
	The notation \( [\,all,all,all\,]_{\mEQ} \) abbreviates
	\( [\,
		\{\forall\,\exists\}^{*}, 
		(\omega)_{i\in\mathbb{N}},
		(\omega)_{i\in\mathbb{N}},	  
	\,]_{\mEQ} \) 
	and simply denotes to the set of all first order sentences in \( \PNF \)
	with an arbitrary signature with equality. 
\end{remark}
\end{definition}

% \begin{example}
% 	The sentence 
% 	\( 
% 		\forall x \exists y \, \mP(\mf(x)\lor\lnot\mQ(y)))
% 		\in	[\, \forall\exists, (\omega, 1), (1) \,] 
% 	\),
% 	i.e.~sentences with prefix 
% 	\( \forall\exists \),
% 	arbitrarily many predicate symbols of arity 1, one predicate symbol of one arity 2, 
% 	no predicate symbol of arity 3 or more,
% 	one function symbol with arity 1,
% 	and no function symbol with arity 2 or more.
% \end{example}

\noindent
Quantifier free first oder logic \([\emptyset, \omega, \omega ]\) is decidable \cite{Kroening:2008:DPA:1391237}, 
but let us take a quick look at two historical non-trivial fragments of first order logic with equality where satisfiability is decidable.

\begin{example}\label{ex:ackermann}
	The Ackermann prefix class contains formulae with arbitrary many existential quantifiers,
	but just one universal quantifier. It contains arbitrary many predicate symbols
	with arbitrarily arities, the equality symbol, but no function symbols.
	\begin{gather*}
	% {\colG {[\,\exists^*\forall\exists^*,all,(1)\,]}_{\mEQ}
	% \quad\supsetneq}\qquad
	{[\,\exists^*\forall\exists^*,all,(0) \,]}_{\mEQ}\qquad\qquad\qquad
	\tag{Ackermann 1928}
	\end{gather*}
\end{example}

\begin{example}\label{ex:monadic}
	The monadic predicate calculus includes sentences with arbitrary quantifier prefixes,
	arbitrarily many unary predicate symbols, the equality symbol, but no function symbols.
	\begin{gather*}
	% {\colG {[\,all, (\omega), (1)\,]}_{\mEQ}
	% \quad\supsetneq}\qquad
	{[\,all, (\omega), (0 ) \,]}_{\mEQ}
	\tag{Löwenheim 1925, Kalmár 1929}
	\end{gather*}
\end{example}

Both examples above are subsets of slightly bigger fragments of first order logic 
with equality where one single unary function symbol can be present
(see Table \ref{tab:decidable:InfinityAxioms}).



% \begin{remark}
% 	One unary function symbol can be added to these fragments of first order logic above
% 	without loosing decidability (see Table~\vref{tab:decidable:InfinityAxioms}).
% \end{remark}

%Goldfarb, Gurevich, Rabin, Shelah completely characterized decidable and undecidable prefix classes.

\begin{table}[hbt]
	\begin{align*}
	% \colG [\,all, (\omega), (0 )&{\colG\,]}_= \tag{Löwenheim 1925, Kalmár 1929}
	% \\
	[ \, \exists^{∗}\forall^{∗}, all, (0)& \, ]{}_{\mEQ} \tag{Bernays, Schönfinkel 1928, Ramsey 1932}
	\\
	[\,\exists^{∗}\forall^2\exists^{∗} , all, (0)&\,] \tag{Gödel 1932, Kalmár 1933, Schütte 1934}
	\\
	[\,all, (\omega), (\omega)&\,] \tag{Löb 1967, Gurevich 1969}
	\\
	[\,\exists^{∗}\forall\exists^{∗}, all, all&\,] \tag{Gurevich 1973}
	\\
	[\,\exists^{∗}, all, all& \, ]{}_{\mEQ} \tag{Gurevich 1976}
	\end{align*}
	\caption[Decidable classes (finite)]{Decidable prefix-vocabulary classes with finite model property}\label{tab:decidedable:FiniteModelProperty}
\end{table}

\begin{definition}
	\coloremph{Axioms of infinity} refers to satisfiable sentences without a finite model.
	\end{definition}

\begin{table}[hbt]
	\begin{align*}
	[ \, all, (\omega), (1)& \, ]{}_{\mEQ} \tag{Rabin 1969}
	\\
	[ \, \exists^{∗}\forall\exists^{∗}, all, (1)& \, ]{}_{\mEQ} \tag{Shelah 1977}
	\end{align*}
	\caption[Decidable classes (infinite)]{Decidable prefix-vocabulary classes with infinity axioms}\label{tab:decidable:InfinityAxioms}
\end{table}

\begin{example}[\cite{MR1482227}]
	Two examples of infinity axioms, each contained 
	in one of the prefix-vocabulary classes from 
	Table~\vref{tab:decidable:InfinityAxioms}.
	\begin{align*}
		\exists x \forall y \forall z\, 
		\left(
			\mf(y) \mNE x \land (\mf(y) \mEQ \mf(z) \limp y \mEQ z) 
		\right)
		&\in [ \, all, (\omega), (1) \, ]{}_{\mEQ} \\
		\forall x\exists y 
		\left(
			\mf^2(y)\mEQ\mf(x) \land \mf(y)\mNE x
		\right)
		&\in [ \, \exists^{∗}\forall\exists^{∗}, all, (1) \, ]{}_{\mEQ}
	\end{align*}
	\end{example}

\begin{lemma}[\cite{MR1482227}]
	Satisfiability is decidable within the prefix-vocabulary classes from
	Tables~\ref{tab:decidedable:FiniteModelProperty} and~\vref{tab:decidable:InfinityAxioms}.
	% Each of theses classes is closed under conjunction with respect to satisfiability.
	% [\cite{MR1482227, p.227}]
\end{lemma}

\noindent We notice that these decidable fragments of first order logic are 
tiny and limited in their applications.



\section{Theories in First Order Logic}\label{sec:decidable:fol:theories}

We follow the definitions and examples of first order theories
by A. Middeldorp in~\cite{AM2015L}.

\begin{definition}[Theory]
	A \coloremph{first-order theory} is a pair of a first-order signature
	and the possibly infinite conjunction \( \bigwedge_i A_i \) of first-order formulae,
	i.e.~the axioms, over the theory's signature.
	A theory is \coloremph{consistent} if the contradiction is not entailed.
	A theory is \coloremph{satisfiable} if there exists a model for its axioms.
%
	A \coloremph{theorem} \( \phi \) is a sentence over the theory's signature,
	i.e.~a closed formula, that holds in any model for the theory's axioms.
	\[
		\bigwedge_i A_i \models \phi
		\quad\text{ or }\quad
		\bigwedge_i A_i \limp \phi
	\]
	A theory is \coloremph{decidable} if there is a decision procedure whether an arbitrary sentence holds in the theory, i.e.~if the sentence is a consequence of the axioms.
\end{definition}
%
\begin{example}
A theory with axioms \( \forall x\,\mP(x) \) and \( \exists x\,\lnot\mP(x) \) is neither consistent nor satisfiable.
\end{example}

\begin{lemma}
	A first order theory is consistent if and only if it is satisfiable.
\end{lemma}
%
Even if we can decide that a sentence is not a theorem in a theory,
e.g.~if we find a model for the axioms where the sentence does not hold,
we still cannot conclude that the negation of the sentence is a theorem.
\begin{example}
	The sentence \( F = \forall x \forall y\, (x\mEQ y) \) is definitely not a theorem in many theories, but its negation \( \lnot F \equiv \exists x \exists y\, (x \mNE y) \) is not a theorem in any theory that allows models with a universe of cardinality one.
\end{example}
\begin{definition}\label{def:complete:theory}
	A (consistent) theory is \coloremph{complete} iff for any sentence in the theory (either) the sentence or its negation is a theorem in the theory.
\end{definition}
%
\begin{remark}
In refutation theorem proving
we show the unsatisfiability
of a negated sentence,
i.e.~a \coloremph{conjecture} \( \psi \),
in conjunction with the axioms
to conclude that the conjecture is indeed a theorem.
\[
	\lnot\left(\bigwedge_i A_i \limp \mathsf{conj}\right) \equiv
	\lnot\left(\lnot\bigwedge_i A_i \lor \mathsf{conj}\right) \equiv
	\bigwedge_i A_i \land \lnot \mathsf{\psi}
\]
\end{remark}

\subsection{Theory of Equality}

The following equivalence and congruence axioms form the theory of equality over a first order signature.

\begin{definition}
	[Equivalence] A binary relation \( \mEQ \) over a domain
	is an equivalence relation if and only if the following axioms hold over the given domain.\label{def:equivalence:axioms}
\begin{align*}
\forall x\,
&\left(x \mEQ x\right)
\tag*{reflexivity}
\\
\forall x \forall y\,
&\left(x \mEQ y \limp y \mEQ x \right)
\tag*{symmetry}
\\
\forall x \forall y \forall z\,
&\left(x \mEQ y \land y\mEQ z \limp x\mEQ z\right)
\tag*{transitivity}
\end{align*}
\end{definition}

\begin{remark}
	% The equivalence axioms are expressible within
	% the decidable first-order fragments
	% \(
	% 	[ \, \forall^3, (0), (0) \, ]{}_{\mEQ}
	% 	\subsetneq
	% 	[ \, all, (\omega), (1) \, ]{}_{\mEQ}
	% \).

	The decidable fragments \( [\,\exists^{∗}\forall^2\exists^{∗} , all, (0) \,] \) 
	and \( [\,\exists^{∗}\forall\exists^{∗}, all, all \,] \)
	from Table \vref{tab:decidedable:FiniteModelProperty}
	contain binary predicate symbols. 
	We could introduce a fresh binary predicate symbol, e.g.~\( \mathsf{equals}\),
	but we would gain nothing, because we cannot express transitivity of \( \mathsf{equals} \)
	within these fragments of first order logic.
\end{remark}


\begin{definition}[\( \vec{x} \)-Notation]
	Occasionally we may abbreviate a sequence of \( n \) variables by \( \vec{x} \).
	Then we write \( \mcf(\vecn{x}) \)
	for first-order expression \( \mcf(x_1,\ldots,x_n) \)
	with \(n\)-ary function or predicate symbol \( \mcf \),
%	 \( \mcf\in\mcFn \),
	a single equation \( \vecn{x}\mEQ\vecn{y} \) for
	the conjunction of \( n \) equations
	\( x_1\mEQ y_1\land\ldots\land x_n\mEQ y_n \),
	and \( \forall\vecn{x} \) for
	the sequence of quantified variables \( \forall x_1\ldots\forall x_n \).
\end{definition}

\begin{definition}
	[Congruence schemata] An equivalence relation \( \mEQ \) is a congruence relation if and only if
	the following formulae hold\label{def:congruence:schemata}:
	\begin{align*}
	\forall\vecn{x}\,\forall\vecn{y}\,
	&\left(\vecn{x}\mEQ\vecn{y} \limp \mf(\vecn{x}) \mEQ \mf(\vecn{y}) \right)
	\tag*{for all \( \mf\in\mcFfn \)}
	\\
	\forall\vecn{x}\,\forall\vecn{y}\,
	&\left(\vecn{x}\mEQ\vecn{y} \limp (\mP(\vecn{x}) \limp \mP(\vecn{y}))\right)
	\tag*{for all \( \mP\in\mcFPn \)}
	\end{align*}
\end{definition}

%\begin{remark}For simplicity we have replaced logical biconditionals in symmetry and predicate congruence with implications.
%	\begin{align*}
%	\forall x \forall y\,
%	\left(x \mEQ y \lbic y \mEQ x \right)
%	&
%	\equiv\forall x \forall y\,
%	\left(x \mEQ y \limp y \mEQ x \right)
%	\\
%	\forall\vec{x}\,\forall\vec{y}\,
%	\left(\vec{x}\mEQ\vec{y} \limp (\mP(\vec{x}) \lbic \mP(\vec{y}))\right)
%	&\equiv
%	\forall\vec{x}\,\forall\vec{y}\,
%	\left(\vec{x}\mEQ\vec{y} \limp (\mP(\vec{x}) \limp \mP(\vec{y}))\right)
%	\end{align*}
%\end{remark}
%
\begin{lemma}\label{lemma:equivalence:congruence}
The equivalence and congruence axioms of equality are provable with natural deduction
(Definition~\vref{def:natural:deduction},
Table~\vref{tab:natural:deduction:connectives},
Table~\vref{tab:natural:deduction:equality},
and Table~\vref{tab:natural:deduction:quantifiers}).
\end{lemma}
\begin{proof}
	For brevity we skip the quantifier introductions (and handle variables like constants)
	for symmetry, transitivity, and congruence.
	Additionally we just show congruence for a unary function and a unary predicate symbol.
	\begin{gather*}
\begin{BMAT}{rcrclccl}{cc}
1 && x_0 && x_0=x_0 	& && {=}i\\
2 && 	 &&	\forall x\,(x=x)	& && \forall{}i,1,\{x\mapsto x_0\}
\addpath{(2,1,1)rrrrulllld}
\end{BMAT}
\\[0.7em]
\begin{BMAT}{rcrclccl}{ccccccc}
1 && && y = y   & && {=}i\\
2 && && x = y	& && \texttt{assume}\\
3 && &&	y\neq x	& && \texttt{assume}\\
4 && && y\neq y & && {=}e,2,3\\
5 && && \bot    & && {\lnot}e,1,4 \\
6 && && y = x   & && \texttt{PBC}, 3{-} 5\\
7 && && x = y \limp y = x & && {\limp}i, 1{-} 5
\addpath{(2,1,1)rrrruuuuullllddddd}
\addpath{(3,2,1)rruuullddd}
\end{BMAT}
\hspace{5em}
\begin{BMAT}{rcrclccl}{ccccccc}
1 && && \mf(y) = \mf(y)   & && {=}i\\
2 && && x = y	& && \texttt{assume}\\
3 && &&	\mf(x) \neq \mf(y)	& && \texttt{assume}\\
4 && && \mf(y) \neq \mf(y) & && {=}e,2,3\\
5 && && \bot    & && {\lnot}e,1,4 \\
6 && && \mf(x) = \mf(y)   & && \texttt{PBC}, 3{-} 5\\
7 && && x = y \limp \mf(x) = \mf(y) & && {\limp}i, 1{-} 5
\addpath{(2,1,1)rrrruuuuullllddddd}
\addpath{(3,2,1)rruuullddd}
\end{BMAT}
\\[0.7em]
\begin{BMAT}{rcrclccl}{ccccc}
1 &&     && x=y\land y=z 	& && \texttt{assume}\\
2 && 	 &&	y=z	& && \land{}e_2, 1 \\
3 && 	 &&	x=y	& && \land{}e_1, 1 \\
4 && 	 &&	x=z	& && {=}e,2,3 \\
5 && 	 &&	x=y\land y=z\limp x=z	& && {\limp}i,2{-}4
\addpath{(2,1,1)rrrruuuulllldddd}
\end{BMAT}
\hspace{3em}
\begin{BMAT}{rcrclccl}{ccccc}
1 && && x = y	& && \texttt{assume}\\
2 && &&	\mP(x)	& && \texttt{assume}\\
3 && && \mP(y) & && {=}e,1,2\\
4 && && \mP(x) \limp \mP(y)   & && {\limp}i, 2{-} 3\\
5 && && x = y \limp (\mP(x)\limp\mP(y)) & && {\limp}i, 1{-} 4
\addpath{(2,1,1)rrrruuuulllldddd}
\addpath{(3,2,1)rruulldd}
\end{BMAT}
\end{gather*}
\end{proof}

\subsection{Natural numbers}

The following axioms characterize natural numbers, addition, and multiplication.

\begin{definition}
	[Natural Numbers]\label{def:natural:numbers:axioms}
	We introduce a constant symbol \(0\),
	a unary successor symbol \( \msucc\in\mcFn[1] \) --- congruence must hold ---
	and restrict the possible models with two axioms.
\begin{align*}
\forall x\, &(\msucc(x) \mNE 0)
\tag*{zero is minimal}
\\ %–––––––––––––––––––––––––––
\forall x \forall y\, &(\msucc(x) \mEQ \msucc(y) \limp x \mEQ y)
\tag*{injectivity of \( \msucc \)}
\\
\colG\forall x\forall y\,&\colG(x\mEQ y \limp \msucc(x)\mEQ\msucc(y))
\tag*{congruence of \( \msucc \)}
\\
\underbrace{G(0)}_{\text{base}}
\land  \, \forall x'\,&
\underbrace{
	\left(G(x') \limp G(\msucc(x'))\right)
}_{\text{step case}}
\, \limp \, \forall x \, G(x)
\tag*{induction schema}
\end{align*}
\end{definition}

\begin{remark}
	The axioms of natural numbers are expressible within the
	decidable first-order fragment
	\(
		[ \, \exists\forall^2, (0), (1) \, ]{}_{\mEQ}
		\subsetneq
		[ \, all, (\omega), (1) \, ]{}_{\mEQ}
	\), e.g.~\( \exists z \forall x (\msucc(x)\mNE z)\).
\end{remark}

\begin{example}
	We may prove \( \forall x\, (\msucc(x) \mNE x) \) with \( G(x) = \msucc(x) \mNE x \) by induction.
	\begin{align*}
		\underbrace{\msucc(0)\mNE 0}_{\text{base}}
		\land \, \forall x' \, &
		\underbrace{
			\left( \msucc(x')\mNE x' \limp \msucc(\msucc(x'))\mNE \msucc(x') \right)
		}_{\text{step case}}
		\, \limp \, \forall x \, \msucc(x)\mNE x
		\end{align*}
\end{example}

\begin{definition}
	[Addition]\label{def:addition:axioms}
	We introduce the binary addition symbol \( +\in\mcFfn[2] \)
	--- congruence must hold ---
	with two axioms about defining equalities of sums.
\begin{align*}
\forall x\,
&(x+0 \mEQ x)
\tag*{addition of zero}\\
\forall x\forall y\,
&(x+\msucc(y)) \mEQ \msucc(x+y)
\tag*{addition of non-zero}
\\
\colG\forall x_1\forall x_2\forall y_1\forall y_2\,
&\colG (x_1\mEQ y_1\land x_2\mEQ y_2 \limp x_1+y_1\mEQ x_2+y_2)
\tag*{congruence of \( + \)}
\end{align*}
\end{definition}

\begin{example}
	\begin{gather*}
	{\colG \msucc(\msucc(\msucc(0)))} + \msucc(\msucc(0))
	\mEQ\msucc({\colG \msucc(\msucc(\msucc(0)))} + \msucc(0))
	\mEQ \msucc(\msucc(
	{\colG \msucc(\msucc(\msucc(0)))} + 0))
	\mEQ \msucc(\msucc(
	{\colG \msucc(\msucc(\msucc(0)))}))
	\end{gather*}
\end{example}

\begin{theorem}Presburger arithmetic (Moj\'{z}esz Presburger, 1929), i.e.\ the first order theory that includes the axioms for equality, natural numbers, addition, and 
	infinitely many induction schemata for all possible formulae, is a consistent, decidable, and complete (as in Definition~\vref{def:complete:theory}) theory. 
	The computational complexity of the decision problem is at least doubly exponential \( 2^{2^{cn}} \)(Fischer and Rabin, 1974), but less than triple exponential (Oppen, 1978, Berman, 1980).
\end{theorem}

\begin{remark}
	Already the axioms of addition (without function congruence)
	are only contained in a non-decidable first order fragment
	\( [ \, \forall^2 , (0), (1,1) \, ]{}_{\mEQ} \). Morover
	sentences in Presburger arithmetic are contained in
	\( [ \, all , (0), (1,1) \, ]{}_{\mEQ} \).
	Still the theory is decidable, even complete.
\end{remark}

\begin{definition}
	[Multiplication]\label{tab:addition:axioms}
	We introduce the binary multiplication symbol \( \times\in\mcFfn[2] \)
	--- congruence must hold ---
	with two axioms about defining equalities of products.
	\begin{align*}
	\forall x\,
	&(x\times 0 \mEQ 0)
	\tag*{multiplication by zero}\\
	\forall x\forall y\,
	&(x\times\msucc(y) \mEQ (x\times y) + x)
	\tag*{multiplication by non-zero}
	\\
	\colG\forall x_1\forall x_2\forall y_1\forall y_2\,
	&\colG (x_1\mEQ y_1\land x_2\mEQ y_2 \limp x_1\times y_1\mEQ x_2 \times y_2)
	\tag*{congruence of \( \times \)}
	\end{align*}
\end{definition}

\begin{theorem}Peano Arithmetic (Giuseppe Peano, 1889),
	i.e.\ the first-order theory that extends Presburger Arithmetic with multiplication, is incomplete (Gödel`s second incompleteness theorem in 1932) and undecidable.
\end{theorem}

\begin{remark}
	Sentences in Peano arithmetic are contained in
	\( [ \, all , (0), (1,2) \, ]{}_{\mEQ} \).
\end{remark}

\begin{theorem}
	The axioms of Peano Arithmetic is consistent (Gentzen, 1936)
	if primitive recursive arithmetic (Skolem, 1923) is consistent.
\end{theorem}

\begin{lemma}[ACN]\label{lem:acn}
	Addition and Multiplication on natural numbers are associative, commutative,
	and determine neutral elements.
	\begin{align*}
	\forall x\forall y\forall z\,&(x\circ(y\circ z) \mEQ (x\circ y)\circ z)
	\tag*{associativity of \( \circ\in \{ +,\times \} \) }
	\\
	\forall x\forall y\,&(x\circ y \mEQ y\circ x)
	\tag*{commutativity of \( \circ\in \{ +, \times \} \)}
	\\
	\forall x\,&(x + 0\mEQ x \land 0 + x \mEQ x )
	\tag*{neutral element for \( + \)}
	\\
	\forall x\,&(x \times \msucc(0)\mEQ x \land \msucc(0) \times x \mEQ x )
	\tag*{neutral element for \( \times \)}
	\end{align*}
\end{lemma}

%\begin{proof}by induction
%	\begin{align*}
%	0 + 0 \mEQ 0
%	\tag*{\( \forall e \)}
%	\\
%	0+x\mEQ x\tag*{assumption}
%	\\
%	0+\msucc(x)\mEQ \msucc(x)
%	\\
%	(0 + 0 \mEQ 0)
%	\land
%	\forall x\,(0+x\mEQ x \limp 0+\msucc(x)\mEQ\msucc(x))
%	\limp
%	\forall y\,(0+y\mEQ y)
%	\tag*{induction}
%	\end{align*}
%\end{proof}










