% !TeX root = ../mthesis.tex
% !TeX encoding = UTF-8
% !TeX spellcheck = en_US

\chapter{Preliminaries}

%\input{epigraphs/defintion}

In this thesis we assume the reader's familiarity with 
propositional and predicate logic \cite{Huth:2004:LCS:975331}, 
%automated theorem proving \cite{Fitting:1996:FLA:230183}, 
term rewriting \cite{Baader:1998:TR:280474}, 
decision procedures \cite{Kroening:2008:DPA:1391237}, 
and satisfiability checking modulo theories \cite{Biere:2009:HSV:1550723}.
Nevertheless -- for clarity -- we state basic notions and  definitions 
of first order logic with equality in section \ref{sec:syntax},
introduce basic concepts of first order semantics in section \ref{sec:semantics},
and describe basic term rewriting terminology in section \ref{sec:termrewriting}. 
%This section is an extension of the same section in our seminar report \cite{axm:SR2}
These notions and notations largely follow those in lecture notes \cite{AM2015tr} and \cite{GM2013ar}.
% Figure \vref{fig:conventions} in the appendix lists typographical conventions.


\section{Syntax}\label{sec:syntax}

In this section we introduce the syntax of arbitrary first order formualae (\FOF) 
and clausal normal form (\CNF).

\input{preliminaries/signature}

\input{preliminaries/terms}

\input{preliminaries/atoms}

\input{preliminaries/ground}

\input{preliminaries/fof}

\input{preliminaries/cnf}

\input{preliminaries/substitution}

\input{preliminaries/unifier}

\input{preliminaries/positions}

\section{Semantics}\label{sec:semantics}

In this section we recall some basic aspects and definitions of semantics in first order logic. 
We state satisfiability of clauses. See the appendix for more details and mathematics.

\subsection{Models}

\begin{definition}\label{def:interpretation}
	An {\myem interpretation} $\mcI$ over a signature $\mcF$ consists of a
	non-empty set $A$ -- the {\myem universe} or {\myem domain},
	definitions of mappings $\mf_\mcI: A^n\rightarrow A$ for every function symbol $\mf\in\mcFf$, 
	and definitions of (possibly empty) n-ary relations 
	 ${\mP_\mcI}\subseteq A^n$ for every predicate symbol $\mP\in\mcFP$
	 and the definition of a binary relation ${\mEQ_\mcI}\subseteq A^2$ for the equality symbol.
	 A (variable) {\myem assignment} is a mapping from variables to elements of the domain.
	 We define the {\myem evaluation} $\alpha_\mcI$ of a term $t$ 
	 for assingment $\alpha$ and interpretation $\mcI$:
	 \DEFINE{\alpha_\mcI(t)}{ 
	 	\alpha_\mcI(x) 
	 	&\text{if }t=x\in\mcV \\
	 	\mc_\mcI
	 	&\text{if }t=\mc\in\mcFfn[0]
	 	\\
	 	\mf_\mcI (\alpha_\mcI(t_1),\ldots,\alpha_\mcI(t_n)) 
	 	&\text{if }t=\mf(t_1,\ldots,t_n), \mf\in\mcFfn[n>0], t_i\in\mcTf 
 }
\begin{remark}
	The evaluation of ground terms does not depend on variable assignments.
\end{remark}
\end{definition}

\begin{definition}\label{def:semantics:atoms}
	A predicate $\mP(t_1,\ldots,t_n)$ 
	{\myem holds} for an assignment $\alpha_\mcI$ 
	if and only if the evaluation of its n-tuple $
	(\alpha_\mcI({t_1}),\ldots,\alpha_\mcI({t_n}))$
	is an element of the relation $\mP_\mcI\subseteq A^n$.
	Similar an equation $s\mEQ t$ 
	holds if $\alpha_\mcI(s) \mEQ_\mcI \alpha_\mcI(t)$.
\end{definition}

\begin{definition}[\index*{Semantics} of \index{FOF}\FOF]\label{def:semantics:FOF}
	A universally quantified formula $\forall x F$ holds in an interpretation if its subformula $F$ holds for all assignments for $x$.
	An existential quantified formula $\exists xF$ holds if its subformula $F$ holds for at least one assignment for $x$.
	A negation $\lnot F$ holds if its subformula $F$ does not hold, 
	a disjunction $F\lor G$ holds if one or both of its subformulae $F$ or $G$ hold,
	a conjunction $F\land G$ holds, if both of its subformualae $F$ and $G$ hold, 
	an implication $F\limp G$ holds if its first subformula $F$ does not hold or its second subformula $F'$ holds (or both).
	\begin{remark}Usually we use precedences on connectives to omit parentheses 
		and some heuristics to structure the formulae for readability 
		without introducing semantic ambiguity.
	\end{remark}
\end{definition}

\begin{definition}[\index*{Semantics} of \index{CNF}\CNF]\label{def:semantics:CNF}
	An atom holds in an interpretation if and only if it 
	holds with all possible assignments. 
	A literal holds if and only if its complement does not hold.
	A clause holds if at least one of its literals holds,
	hence the empty clause $\square$ does not hold in any interpretation.
	A set of clauses holds if and only if every clause in the set holds.
\end{definition}

\begin{definition}
	A {\myem model} $\mcM$ for a set of clauses (a formula) is an interpretation that 
	{\myem satisfies} the set of clauses (the formula),
	i.e.~the set of clauses (the formula) holds in this interpretation.
	A set of clauses (a formula) is {\myem satisfiable} if and only if there is at least one model for it. 
	A set of clauses (a formula) is {\myem valid} if and only if every interpretation is a model.
\end{definition}

\begin{example}
	The set of clauses $\{\,\lnot\mE(\mzero, \mf(x))\,\}$ is satisfiable if and only if the quantified first order formula $\exists x \forall y\,(\lnot \mE(x,y))$ without free variables is satisfiable.
\end{example}

\begin{definition}\label{def:hk}
	The {\myem Herbrand universe} for a first order signature $\mcF$ 
	is the smallest set of terms that contains all $H_{i\geq 0}$ defined inductively as
	\begin{align*}
	H_0 &= \left\{ 
	\begin{array}{ll}
	\{\,\mc \mid \mc\in\mcFfn[0]\,\} 
	&\text{if } \mcFfn[0]\not=\emptyset\\
	\{\,\mc_0\,\}
	&\text{if } \mcFfn[0]=\emptyset, \mc_0\not\in\mcF
	\end{array}
	\right. 
	&
	H'_0 &= H_0
	\\
	H_{k+1} &= \bigcup_{n> 0} 
	\{\,  
	\mf(t_1,\ldots,t_n) \mid
	\mf\in\mcFfn[n],
	t_1,\ldots,t_n \in H_k'
	\,\}
	&
	H'_{k+1} &= H_k \cup H'_{k+1}
	\end{align*}
	
\end{definition}

\begin{definition}
	An {\myem Herbrand interpretation} $\mcH$ is an interpretation where the domain 
	is an Hebrand universe
	and the interpretation of each ground term $t_\mcH := t$ is the term itself.
\end{definition}

\subsection{Equality}

In Definition \ref{def:semantics:atoms} we have interpreted the equality symbol as binary relation without restrictions.
This of course allows undesirable models as in the following Example \ref{ex:a:neq:a}.
Hence we introduce useful definitions to deal with this situation and demonstrate their usage in an example. 

\begin{example}
	\label{ex:a:neq:a}
	Any interpretation $\mcI$  
	with ${\mEQ_\mcI} = \emptyset$ satisfies $\ma\mNE\ma$.
\end{example}

\begin{definition}\label{def:normal:interpreation}
	An {\myem normal} interpretation defines $\mEQ_\mcI$ as identity on its domain,
	e.g.~the equation of terms $s \mEQ_\mcI t$ holds if and only 
	if any evaluation of its terms are equal $\alpha_\mcI(s) = \alpha_\mcI(t)$ 
	for all assignments $\alpha$. 
	In other words a normal interpretation yields different elements 
	for ground terms $s'$ and $t'$ if and only if $s'\mNE_\mcI t'$.
\end{definition}

\begin{definition}
	A {\myem term interpretation} 
	$\mcI_t$ 
	is an interpretation 
	where the elements of its domain $A = \mcTFf/_\sim$ 
	are equivalence classes of ground terms
	and the interpretation of each ground term $t^{\mcI_t} := [t]_\sim$ is its equivalence class.
	%		An equation $s\mEQ t$ of ground terms holds in if $[s]_\sim=[t]_\sim$.
	A ground predicate $\mP(t_1,\ldots,t_n)$ holds if 
	$([t_1]_\sim,\ldots,[t_n]_\sim) \in \mP^{\mcI_t} \subseteq A^n$.
	%	A ground literal does not hold if and only if its complementary literal holds.
	%	In an {\myem equational} term interpretation an equation $s\mEQ t$ holds if an only if $s\sim t$.
\end{definition}


\begin{example}
	Consider the satisfiable set of clauses $S = \{ \mf(x) \mEQ x \}$. 
	%	
	We easily find a Herbrand model $\mcH$ with
	predicate definition $\mEQ_\mcH = \{ (\mf^{i+1}(\ma)), \mf^{i}(\ma)) \mid i\geq 0  \} $. 
	However $\mcH$ is not a normal model because obviously $\mf(\ma)\neq\ma$ in its domain.
	%	
	Further on we easily find an normal model $\mcM$ 
	with domain $\{ \mc \}$, function definition $\mf_\mcM(\mc) \mapsto \mc$, 
	and the relation ${\mEQ_\mcM} = \{ (\mc,\mc)\}$ coincides with identity in its domain.
	Certainly this model $\mcM$ is not an Hebrand model 
	because the interpretation of ground term ${\mf(\mc)}_\mcM = \mc \neq \mf(\mc)$ is not the term itself.
	%	
	On the other hand we easily get a normal term model $\mcM_t$ with domain $\{ [\ma]_\sim\}$, 
	a plain function definition
	$\mf_{\mcM_t}([\ma]_\sim) \mapsto [\ma]_\sim$ 
	with equivalence relation
	$\ma \sim \mf(\ma)$.
	Hence $\mEQ_{\mcM_t}$ agrees to equality in its domain.	
\end{example}

How we find a normal model or justify its existence we will discuss in Section \vref{sec:automation} about Automation.


\section{Term Rewriting}\label{sec:termrewriting}


\begin{definition}\label{def:rewriterule}
	A {\myem rewrite rule} is an equation of terms where the left-hand side is not a variable
	and the variables occuring in the right-hand side occur also in the left-hand side.
	%	\[
	%		\ell\rwEQ r \text{ is rewrite rule }\quad :\Longleftrightarrow\quad\ell\not\in\mcV\text{ and }\var(r)\subseteq\var(l)
	%	\] 
	A rewrite rule $\ell'\rwStep r'$ is a {\myem variant} of $\ell\rwStep r$ if there is a renaming $\varrho$
	such that 
	$(\ell\rightarrow r)\varrho := \ell\varrho\rightarrow r\varrho = l'\rightarrow r'$.
	A {\myem term rewrite system} is a set of rewrite rules without variants.
\end{definition}
%
\begin{definition}
	We say $s\rightarrow_\mcR t$ is a 
	{\myem rewrite step} 
	 with respect to TRS $\mcR$ 
	when there is a position $p \in \pos(s)$, 
	a rewrite rule $\ell\rwStep r\in\mcR$, 
	and a substitution $\sigma$ such that
	$s|_p=\ell\sigma$ and $s[r\sigma]_p = t$.
		The subterm $\ell\sigma$ is called {\myem redex} and
	$s$ rewrites to $t$ by {\myem contracting} $\ell\sigma$ to {\myem contractum} $r\sigma$.
	%
	We say a term $s$ is {\myem irreducible} or in {\myem normal form} with respect to TRS $\mcR$ if there is no rewrite step $s\rightarrow_\mcR t$ for any term $t$. 
	The set of normal forms $\mNFR$ contains all irreducible terms of the TRS $\mcR$.
	\end{definition}
%
%\begin{definition}
%	A {\myem rewrite relation} is a binary relation on terms that is closed under contexts and substitutions.
%	A {\myem rewrite order} is a proper order on terms
%	(i.e. an irreflexive and transitive relation on terms) and a rewrite relation.
%	A {\myem reduction order} is a well-founded rewrite order.
%\end{definition}
%
\begin{definition}
	A term $s$ can be rewritten to term $t$ with notion $s\rightarrow^*_\mcR t$ 
	if there exists at least one {\myem rewrite sequence} $(a_1,\ldots ,a_n)$ such that
	$s=a_1$, $a_n=t$, and $a_i\rightarrow_\mcR a_{i+1}$ are rewrite steps for $1\leq i<n$.
	A TRS is {\myem terminating} if there is no infinite rewrite sequence of terms.
	%
	Two Terms $s$ and $t$ are {\myem joinable} with notion $s\downarrow t$ 
	if both can be rewritten to some term $c$, i.e.~$s \rightarrow^*c\ \, ^*\!\!\leftarrow t$.
%	
	Two Terms $s$ and $t$ are {\myem meetable} with notion $s\uparrow t$ 
	if both can be rewritten from some common ancestor term $a$, i.e.~$s \leftarrow^*a\ \, ^*\!\!\rightarrow t$.
%
	A TRS is {\myem confluent } if $s$ and $t$ are joinable whenever $s\ ^*\!\!\leftarrow a \rightarrow^* t$ holds for some term $a$.
	%
	Terms $s$ and $t$ are {\myem convertible} with notion $s\leftrightarrow^* t$ 
	if there exists a sequence $(a_1,\ldots ,a_n)$ such that
	$s=a_1$, $a_n=t$, and $a_i\leftrightarrow a_{i+1}$, i.e.~$a_i\rightarrow a_{i+1}$ or $a_i\leftarrow a_{i+1}$ are rewrite steps for $1\leq i<n$.
\end{definition}
%
\begin{definition}\label{def:closed-under}
	A {\myem rewrite relation} is a binary relation 
	$\relation$ on arbitrary terms $s$ and $t$, 
	which additionally is {\myem closed under contexts} 
	(whenever $s\relation t$ then $u[s]_p\relation u[t]_p$ 
	for an arbitrary term $u$ and any position $p\in\pos(u)$)
	and {\myem closed under substitutions} 
	(whenever $s\relation t$ then $s\sigma\relation t\sigma$
	for an arbitrary substitution $\sigma$).
\end{definition}
\begin{lemma}
	The relations $\rightarrow^*_\mcR$, 
	$\rightarrow^+_\mcR$,
	$\downarrow_\mcR$, $\uparrow_\mcR$ are rewrite relations on every TRS $\mcR$.
\end{lemma}
%%
\begin{definition}
	A proper (i.e.~irreflexive and transitive) order on terms is called {\myem rewrite order} if it is a rewrite relation.
	A {\myem reduction order} is a well-founded rewrite order,
	i.e.~there is no infinite sequence 
	$(a_i)_{i\in\mathbb{N}}$
	where $a_i\succ a_{i+1}$ for all $i$.
	% with $i\in\mathbb{N}$.
	A {\myem simplification order} is a rewrite order with the {\myem subterm property},
	i.e.~$u[t]_p \succ t$ for all terms $u$, $t$ and positions $p\neq\epsilon$.
\end{definition}
\begin{lemma}
	Every simplification order is well-founded, hence it is a reduction order.
\end{lemma}
%%% DEF %%
%%
%\begin{theorem}
%	A TRS $\mcR$ is terminating if and only if there exists a reduction order $\gtpre$
%	such that $l\gtpre r$ for every rewrite rule $l\rightarrow r\in\mcR$.
%	We call $\mcR$ simply terminating if $\gtpre$ is a simplification order.
%\end{theorem}



